Kids First or Marriage First?

### 2019-05-31 main/model.tex
Received fixes from Martí

Couples with kids have additional choice variable $q$, that can be interpreted as flow of child quality or child consumption. Child quality is produced using mother's labor $l_f$ and monetary expenditures $x$ according to constant returns to scale production function. Labor input to child quality causes female to lose part of her labor income. The production function depends on age, namely I parametrize it as $q = [\mu_\xi\cdot l_f^{\theta_q} + (1-\mu_\xi)\cdot x^{\theta_q}]^{\frac1{\theta_q}}$. This follows, for instance, [Sommer ...], with the addition of changing labor share $\mu$.

\begin{align}& \hspace{5em}  V^{ck}_t(\Omega) = \max\limits_{c^f,c^m,q,l_f,x}  \bigg\{ \theta\cdot u(c^f,q) + (1-\theta)\cdot u(c^m,q) + \psi + \label{vf_ck} \\  \nonumber
 & \hspace{10em} \beta \E_t \Big[   \left\{ (1-d)\cdot   V^{ck}_{t+1}(\Omega') + d\cdot [ \theta V_{t+1}^{f,sk}(\omega^{df}) + (1-\theta)V_{t+1}^{m,s}(\omega^{dm})]\right\} \Big] \bigg\},
\end{align}\vspace{-2em}
\begin{align*}
\text{s.t. \ } & a' + c + x = R\cdot a  + W^m_t(z^m) + (1-l_f)\cdot W^f_t(z^f) & \text{(evolution of joint assets)},\\
                    & c = [(c^f)^{1+\rho_c} + (c^m)^{1+\rho_c}]^{\frac1{1+\rho_c}} & \text{(increasing returns in consumption)},\\
                    & q = [\mu_\xi\cdot l_f^{\theta_q} + (1-\mu_\xi)\cdot x^{\theta_q}]^{\frac1{\theta_q}} & \text{(production function of child quality)},\\
                    &  z^{f\prime} = z^f + \varepsilon^{z,f}_t, \ \ \varepsilon^{z,f}_t \sim \mathcal{N}(0;\sigma_{z,f}^2) &  \text{ (evolution of female productivity)}\\
				 &  z^{m\prime} = z^m + \varepsilon^{z,m}_t, \ \ \varepsilon^{z,m}_t \sim \mathcal{N}(0;\sigma_{z,m}^2) &  \text{ (evolution of male productivity)}\\
                    & \psi' = \psi + \varepsilon^{\psi}_t, \ \ \varepsilon^{\psi}_t \sim \mathcal{N}(0;\sigma_{\psi}^2)  & \text{(evolution of marriage surplus),} \\
                   &  \P(\xi' = s | \xi = y) = p^s, \ \ \P(\xi' = s | \xi = s) = 1 & \text{(evolution of child's age)},\\
                    & (\Omega',\omega^{df},\omega^{dm}) = \mathcal{R}(\theta,a',z^{m\prime},z^{f\prime},\psi',\xi) & \text{(renegotiation correspondence)},
\end{align*}

@@

The problem of a female who stays single mother is
\begin{align}V^{f,sk}_t(\omega) = \max\limits_{c} & \bigg\{ u(c,q) + \beta \E_t \Big[ p^{\text{out}} \cdot \left\{ V^{f,s}_{t+1}(\omega') + S\right\} +(1-p^{\text{out}})V^{f,sk}_{t+1}(\omega') \Big]  \bigg\},
\end{align}\vspace{-1em}
\begin{align*}
 \text{s.t. \ }  &  a' = R\cdot a  + (1-l_f)W^f_t(z) - c  & \text{ (evolution of assets)}\\
 & q = [\mu_\xi\cdot l_f^{\theta_q} + (1-\mu_\xi)\cdot x^{\theta_q}]^{\frac1{\theta_q}} & \text{(production function of child quality)},\\
 &  z' = z + \varepsilon^{z,f}_t, \ \ \varepsilon^{z,f}_t \sim \mathcal{N}(0;\sigma_{z,f}^2) &  \text{ (evolution of productivity)}\\
 &  \P(\xi' = s | \xi = y) = p^s, \ \ \P(\xi' = s | \xi = s) = 1 & \text{(evolution of child's age)},\\
\end{align*}

@@

\subsection{Functional Forms}
I use standard CES utility functions for individuals and couples with no kids: $u(c) = \frac{c^{1-\sigma}}{1-\sigma}$.  When kids appear, people extract additional utility of child quality, I use the following non-separable utility function:
\[u(c,q) = \frac{\left(e^{\phi_0}\cdot c^{1-\phi_1} \cdot (\phi_2 + q)^{\phi_1}\right)^{1-\sigma}}{1-\sigma},\]
parameter $\phi_0$ represents fixed gain in marginal utility because of having a baby, $\phi_1$ drives relative importance of child quality in utility function and $\phi_2$ is a ``luxury good'' parameter that can rationalize large child expenditures of wealthy couples that are discussed in the estimation section.

>>>>

Couples with kids have additional choice variable $q$, that can be interpreted as flow of child quality or child consumption. Child quality is produced using mother's labor $l_f$ and monetary expenditures $x$ according to constant returns to scale production function. Labor input to child quality causes female to lose part of her labor income. The production function depends on age, namely I parametrize it as $q = [\mu_\xi\cdot l_f^{\theta_q} + (1-\mu_\xi)\cdot x^{\theta_q}]^{\frac1{\theta_q}}$. This follows Sommer, 2014, with the addition of changing labor share $\mu$. Also, following the same paper, I introduce the lower bound for possible choices of $q$ (that may rationalize additional distress in case of unplanned pregnancies).

\begin{align}& \hspace{5em}  V^{ck}_t(\Omega) = \max\limits_{c^f,c^m,q,l_f,x}  \bigg\{ \theta\cdot u(c^f,q) + (1-\theta)\cdot u(c^m,q) + \psi + \label{vf_ck} \\  \nonumber
 & \hspace{10em} \beta \E_t \Big[   \left\{ (1-d)\cdot   V^{ck}_{t+1}(\Omega') + d\cdot [ \theta V_{t+1}^{f,sk}(\omega^{df}) + (1-\theta)V_{t+1}^{m,s}(\omega^{dm})]\right\} \Big] \bigg\},
\end{align}\vspace{-2em}
\begin{align*}
\text{s.t. \ } & a' + c + x = R\cdot a  + W^m_t(z^m) + (1-l_f)\cdot W^f_t(z^f) & \text{(evolution of joint assets)},\\
                    & c = [(c^f)^{1+\rho_c} + (c^m)^{1+\rho_c}]^{\frac1{1+\rho_c}} & \text{(increasing returns in consumption)},\\
                    & q = [\mu_\xi\cdot l_f^{\theta_q} + (1-\mu_\xi)\cdot x^{\theta_q}]^{\frac1{\theta_q}} & \text{(production function of child quality)},\\
                    & q \geq \underline{q} & \text{(required childcare costs)},\\
                    &  z^{f\prime} = z^f + \varepsilon^{z,f}_t, \ \ \varepsilon^{z,f}_t \sim \mathcal{N}(0;\sigma_{z,f}^2) &  \text{ (evolution of female productivity)}\\
				 &  z^{m\prime} = z^m + \varepsilon^{z,m}_t, \ \ \varepsilon^{z,m}_t \sim \mathcal{N}(0;\sigma_{z,m}^2) &  \text{ (evolution of male productivity)}\\
                    & \psi' = \psi + \varepsilon^{\psi}_t, \ \ \varepsilon^{\psi}_t \sim \mathcal{N}(0;\sigma_{\psi}^2)  & \text{(evolution of marriage surplus),} \\
                   &  \P(\xi' = s | \xi = y) = p^s, \ \ \P(\xi' = s | \xi = s) = 1 & \text{(evolution of child's age)},\\
                    & (\Omega',\omega^{df},\omega^{dm}) = \mathcal{R}(\theta,a',z^{m\prime},z^{f\prime},\psi',\xi) & \text{(renegotiation correspondence)},
\end{align*}

@@

The problem of a female who stays single mother is
\begin{align}V^{f,sk}_t(\omega) = \max\limits_{c} & \bigg\{ u(c,q) + \beta \E_t \Big[ p^{\text{out}} \cdot \left\{ V^{f,s}_{t+1}(\omega') + S\right\} +(1-p^{\text{out}})V^{f,sk}_{t+1}(\omega') \Big]  \bigg\},
\end{align}\vspace{-1em}
\begin{align*}
 \text{s.t. \ }  &  a' = R\cdot a  + (1-l_f)W^f_t(z) - c  & \text{ (evolution of assets)}\\
 & q = [\mu_\xi\cdot l_f^{\theta_q} + (1-\mu_\xi)\cdot x^{\theta_q}]^{\frac1{\theta_q}} & \text{(production function of child quality)},\\
& q \geq \underline{q} & \text{(required childcare costs)},\\
 &  z' = z + \varepsilon^{z,f}_t, \ \ \varepsilon^{z,f}_t \sim \mathcal{N}(0;\sigma_{z,f}^2) &  \text{ (evolution of productivity)}\\
 &  \P(\xi' = s | \xi = y) = p^s, \ \ \P(\xi' = s | \xi = s) = 1 & \text{(evolution of child's age)},\\
\end{align*}

@@

\subsection{Functional Forms}
I use standard CES utility functions for individuals and couples with no kids: $u(c) = \frac{c^{1-\sigma}}{1-\sigma}$.  When kids appear, people extract additional utility of child quality, I use the following non-separable utility function:
\[u(c,q) = \frac{\left(e^{\phi_0}\cdot c^{1-\phi_1} \cdot (\phi_2 + q)^{\phi_1}\right)^{1-\sigma}}{1-\sigma} + \psi_3,\]
parameter $\phi_0$ represents fixed gain in marginal utility because of having a baby, $\phi_1$ drives relative importance of child quality in utility function and $\phi_2$ is a ``luxury good'' parameter that can rationalize large child expenditures of wealthy couples that are discussed in the estimation section, $\phi_3$ is an additional utility shifter.
